% Chapter 8: Mobile Dashboard
\section{Cross-Platform Mobile Health Dashboard}

\subsection{Technology Stack}
React Native 0.81.5 + Expo SDK 54 + TypeScript. State: Zustand. Navigation: React Navigation 7 (Bottom Tabs). Storage: AsyncStorage (offline cache). Notifications: Expo Notifications. HTTP: Axios. Icons: Expo Vector Icons. Animations: Lottie + react-native-reanimated.

\subsection{Three-Screen Architecture}
\textbf{HomeScreen:} Real-time metrics (MetricCards for HR, steps, calories), anomaly alert cards with human-readable reasons, activity classification display with icons. \textbf{HistoryScreen:} Temporal trends grouped by date, activity + anomaly badges, anomaly reason tooltips. \textbf{SettingsScreen:} Sync interval configuration, notification preferences, manual data entry.

\subsection{Offline-First Strategy}
Data fetching priority: AsyncStorage cache $\rightarrow$ API fetch $\rightarrow$ mock data fallback. Ensures dashboard remains functional during connectivity gaps.

\subsection{Push Notification Integration}
On startup: Expo generates push token $\rightarrow$ POST to \texttt{/notifications/register} $\rightarrow$ stored in DynamoDB. On anomaly: cloud sends push via Expo API with anomaly reasons in the notification body.

\subsection{Performance Results}

\begin{table}[H]
\centering
\caption{Mobile Dashboard Performance}
\label{tab:mobile_perf}
\begin{tabular}{lcc}
\toprule
\textbf{Metric} & \textbf{Measured} & \textbf{Target} \\
\midrule
Initial render (cold) & 480ms & $<$500ms \\
Initial render (warm) & 120ms & $<$200ms \\
Scrolling FPS & 60fps & 60fps \\
API fetch (p50) & 85ms & $<$200ms \\
Memory usage (peak) & 78MB & $<$150MB \\
Bundle size & 12MB & $<$25MB \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Flutter to React Native Migration}
The project migrated from Flutter (Dart) to React Native (TypeScript) for better ecosystem integration. Outcomes: 42 files, 2,500+ LOC, full feature parity with faster build times (30s vs 45s) and simpler state management (Zustand vs Provider).

%% ============================================================
% Chapter 9: Privacy and Security
\section{Privacy-Preserving Edge-First Architecture}

\subsection{Edge-First Privacy Design}
Primary ML inference occurs on-device; only aggregated metrics + ML-derived insights are transmitted to the cloud. Raw continuous sensor streams never leave the watch. Periodic snapshots (every 15--60 min) contain HR values, step counts, edge activity labels, and anomaly scores---no raw accelerometer data, no GPS, no continuous streams.

\subsection{Security Controls}
\textbf{Transit:} TLS 1.2+ (HTTPS-only). \textbf{Authentication:} API key via X-API-Key header at API Gateway. \textbf{Storage:} DynamoDB server-side encryption (AES-256, AWS KMS), S3 bucket-level encryption. \textbf{Access Control:} IAM least-privilege policies per Lambda function.

\subsection{Regulatory Compliance}

\begin{table}[H]
\centering
\caption{Regulatory Compliance Status}
\label{tab:compliance}
\begin{tabular}{llcc}
\toprule
\textbf{Regulation} & \textbf{Requirement} & \textbf{Status} & \textbf{Gap} \\
\midrule
HIPAA & Encryption at rest & Met & None \\
HIPAA & Access controls & Met & Audit logs \\
GDPR & Data minimization & Met & None \\
GDPR & Right to erasure & Partial & No API \\
DPDPA & Data localization & Met & ap-south-2 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Privacy Comparison}

\begin{table}[H]
\centering
\caption{Privacy Comparison with Commercial Systems}
\label{tab:privacy}
\begin{tabular}{lcccc}
\toprule
\textbf{Feature} & \textbf{Apple} & \textbf{Fitbit} & \textbf{Samsung} & \textbf{Ours} \\
\midrule
Edge processing & Partial & No & Partial & \textbf{Yes} \\
Data minimization & Med & Low & Med & \textbf{High} \\
Open-source & No & No & No & \textbf{Yes} \\
Self-hostable & No & No & No & \textbf{Yes} \\
Auditability & Low & Low & Low & \textbf{Full} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Proposed Enhancements}
Future work includes: end-to-end encryption (AES-256-GCM at application level), differential privacy for aggregated statistics ($\epsilon$-differential privacy with Laplace mechanism), and federated learning for collaborative model training without sharing raw data \cite{federated, dp}.
